---
title: Evil-ly delicious spreadsheet
author: Duncan Garmonsway
date: '2017-07-26'
slug: evil-ly-delicious-spreadsheet
categories: ['R', 'tidyxl']
tags: ['R', 'tidyxl']
---

When Bob Rudis needs to import a spreadsheet, he doesn't use R, he uses Twitter.
And when he uses Twitter, I use [tidyxl](https://nacnudus.github.io/tidyxl/).

`r blogdown::shortcode("tweet", "890200287356620806")`

I haven't blogged about [tidyxl](https://nacnudus.github.io/tidyxl/) before,
only tweeted and
[vignetted](https://nacnudus.github.io/unpivotr/articles/small-multiples.html),
so to see what it can really do, follow those links.

The first problem with the spreadsheet is that it lies.  When you
[download](https://t.co/xkfPqROO9E) its file suffix is `xls`, suggesting that it
is in the pre-2007 Excel binary format.  But open it in a text editor, and it
turns out to be xml.

```xml
<?xml version="1.0" encoding="utf-8"?>
<?mso-application  progid="Excel.Sheet" ?>
<Workbook xmlns="urn:schemas-microsoft-com:office:spreadsheet">
  <DocumentProperties xmlns="urn:schemas-microsoft-com:office:office">
    <Author>InstantAtlas™ Server</Author>
  </DocumentProperties>
  <ExcelWorkbook xmlns="urn:schemas-microsoft-com:office:excel">
    <FirstVisibleSheet>0</FirstVisibleSheet>
    <ActiveSheet>0</ActiveSheet>
  </ExcelWorkbook>
  <Styles>
  ...
```

Isn't that the post-2007 `xlsx` Excel file format then?  No.  It's a thoroughly
non-standard hack of it.  Instead of being several xml files in a folder
structure, it's a single file.  Then, when you dig into the xml, it doesn't
define the data in normal cells, but in unusual structure called a 'Table'.

The upshot is that you can't open it with any tool other than Excel.  I tried
all the usual suspects, [readxl](http://readxl.tidyverse.org/),
[openxlsx](https://github.com/awalker89/openxlsx),
[xlsx](https://cran.r-project.org/web/packages/xlsx/index.html),
[XLConnect](https://cran.r-project.org/web/packages/XLConnect/index.html), even
[Office
Online](https://products.office.com/en-gb/office-online/documents-spreadsheets-presentations-office-online) from Microsoft itself.

So the first step is to open it in real, desktop Excel and save it in the xlsx
format.  The rest is comparatively easy.

The next step with any ugly spreadsheet is to do as [David
Robinson](http://varianceexplained.org/about/)
[recommends](https://rpubs.com/dgrtwo/tidying-enron), and 'melt' it into
one-cell-per-row, with three fields giving the row number, column number, and
value of the cell.

You have four options to melt it, because the spreadsheet is simple and all the
cell values are strings.

* Write your own code
* Use David Robinson's [code](https://rpubs.com/dgrtwo/tidying-enron)
* Use the `tidy_table()` function in my
  [unpivotr](https://nacnudus.github.io/unpivotr/reference/tidy_table.html)
  package
* Use the `tidy_xlsx()` function in my
  [tidyxl](https://nacnudus.github.io/tidyxl/reference/tidy_xlsx.html) package.

[Tidyxl](https://nacnudus.github.io/tidyxl/reference/tidy_xlsx.html) really
shines for spreadsheets with more diverse data values and formatting.
[Unpivotr](https://nacnudus.github.io/unpivotr/reference/tidy_table.html) is
great for [small
multiples](https://nacnudus.github.io/unpivotr/articles/small-multiples.html),
does the right thing when
[melting](file:///home/nacnudus/R/unpivotr/docs/reference/tidy_table.html)
tables with different column types, and handles really weird [HTML
tables](file:///home/nacnudus/R/unpivotr/docs/articles/html.html).

```{r, include = FALSE}
library(tidyxl)
library(tidyverse)
library(stringr)
library(here) # a wonderful package
```

```r
library(tidyxl)
library(tidyverse)
library(stringr)
library(here) # a wonderful package
```

```{r, echo = TRUE}
file_path <- file.path(here(), "data", "dataView155_3.xlsx")

# Option 4 (tidyxl)
datasheet <-
  tidy_xlsx(file_path)$data[["SVV Coverage Trend Data"]] %>%
  select(row, col, character)

# Option 3 (unpivotr)
datasheet <-
  readxl::read_excel(file_path, col_names = FALSE) %>%
  unpivotr::tidy_table() %>%
  select(row, col, character = chr)

datasheet
```

Then the worksheet needs to be chopped up into its components: column headers,
row headers, and data cells.  This is where 'melting' the cells pays off, by
making filtering so easy.

```{r, echo = TRUE}
datacells <-
  datasheet %>%
  filter(row >= 4, col >= 2) %>%
  rename(value = character)
state_or_stat <-
  datasheet %>%
  filter(row >= 4, col == 1) %>%
  select(row, state_or_stat = character)
type <-
  datasheet %>%
  filter(row == 1, col >= 2, !is.na(character)) %>%
  select(col, type = character)
header <-
  datasheet %>%
  filter(row == 3, col >= 2) %>%
  select(col, header = character)
year <-
  header %>%
  filter(str_detect(header, "[0-9]{4}-[0-9]{2}")) %>%
  rename(year = header)
```

The separated components look like these two examples.

```{r, echo = TRUE}
datacells
header
```

Finally, join (literally) the components back together.  There is a bit more
cleaning up to do, with standard tidyverse tools.

```{r, echo = TRUE}
output <-
  datacells %>%
  arrange(row, col) %>%
  left_join(state_or_stat, by = "row") %>%
  left_join(year, by = "col") %>%
  left_join(header, by = "col") %>%
  left_join(type, by = "col") %>%
  mutate(header = str_replace(header,
                              "[0-9]{4}-[0-9]{2}",
                              "PERCENT VACCINATED"),
         value = if_else(value %in% c("NA", "NaN", "NReq"), NA_character_, value)) %>%
  fill(year, type) %>%
  select(-row, -col) %>%
  spread(header, value) %>%
  mutate_all(~ parse_guess(.x)) %>%
  arrange(state_or_stat, type, year)

glimpse(output)
```

Someone really keen would join the `FOOTNOTES` column to the footnotes in the
second sheet of the workbook, but that spreadsheet is entirely
un-machine-readable, and has at least one chunk missing.

> \* The percentage of children enrolled in kindergarten surveyed may be
